---
title: "Adding graphical elements"
author: "Mikko Vihtakari (Institute of Marine Research)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Adding graphical elements}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
```

```{r setup}
library(ggOceanMaps)
library(ggplot2)
```

`basemap()` returns a regular `ggplot2` object, so points, paths,
polygons, labels, and custom scales are added with `+`. The only detail
that needs care is the coordinate system. In northern and southern high
latitudes, and whenever `crs` is set, the map axes are projected metres
instead of longitude and latitude. Transform geographic coordinates with
`transform_coord()` before adding them to those maps.

# Schematic current arrows

The
[Barents-Sea-currents](https://github.com/MikkoVihtakari/Barents-Sea-currents)
repository provides publication-style "Figure 1" current arrows for the
Barents Sea, North Atlantic, and Arctic Ocean. The tidy CSV files are easy
to download directly from GitHub and plot on a ggOceanMaps basemap.

```{r barents-current-data}
barents_url <- paste0(
  "https://raw.githubusercontent.com/MikkoVihtakari/",
  "Barents-Sea-currents/master/tabular/barents_currents.csv"
)
barents_file <- file.path(tempdir(), "barents_currents.csv")

if (!file.exists(barents_file)) {
  try(
    download.file(barents_url, barents_file, quiet = TRUE, mode = "wb"),
    silent = TRUE
  )
}

# Offline fallback for building this vignette.
if (!file.exists(barents_file)) {
  local_barents_file <- file.path("data", "barents_currents.csv")
  if (!file.exists(local_barents_file)) {
    local_barents_file <- file.path("vignettes", "data", "barents_currents.csv")
  }
  barents_file <- local_barents_file
}

cur <- read.csv(barents_file)
head(cur)
```

The CSV is stored in decimal degrees. Reproject the arrow nodes to the
map CRS first, then smooth each arrow in projected coordinates. Smoothing
in longitude/latitude can create exaggerated curves near the pole.

```{r barents-current-plot, fig.height = 5.2}
cur <- cur[order(cur$group, cur$order), ]
cur <- transform_coord(
  cur, lon = "long", lat = "lat", proj.out = 32633, bind = TRUE
)

smooth_current <- function(d, shape = -0.6, n = 200) {
  d <- d[order(d$order), ]
  xs <- graphics::xspline(
    d$lon.proj, d$lat.proj,
    shape = shape, draw = FALSE, repEnds = TRUE
  )
  keep <- round(seq(1, length(xs$x), length.out = min(n, length(xs$x))))
  data.frame(
    x = xs$x[keep],
    y = xs$y[keep],
    group = d$group[1],
    type = d$type[1],
    size = d$size[1]
  )
}

grDevices::pdf(NULL)
graphics::plot.new()
cur_smooth <- do.call(rbind, lapply(split(cur, cur$group), smooth_current))
invisible(grDevices::dev.off())

current_cols <- c(Atlantic = "#d7301f", Arctic = "#2166ac")
current_arrow <- arrow(
  type = "open", angle = 15, ends = "last", length = unit(0.25, "lines")
)

reorder_layers(
  basemap(
    limits = c(5, 45, 68, 83.5),
    crs = 32633,
    bathy.style = "rbg",
    land.col = "grey86",
    legends = c(FALSE, TRUE)
  ) +
    geom_path(
      data = cur_smooth,
      aes(x = x, y = y, group = group, colour = type, linewidth = size),
      arrow = current_arrow
    ) +
    scale_colour_manual(
      name = "Current type",
      values = current_cols,
      labels = c(Atlantic = "Atlantic (warm)", Arctic = "Arctic (cold)")
    ) +
    scale_linewidth("Relative strength", range = c(0.3, 1.3))
)
```

# Velocity fields

For a gridded velocity field, calculate arrow endpoints, transform both
the start and end points to the basemap CRS, and draw the result with
`geom_segment()`. The example below uses a small synthetic field filtered
to ocean positions; the same pattern applies after reading `u` and `v`
from a NetCDF file.

```{r quiver-projected, fig.height = 4.5}
grd <- expand.grid(
  lon = seq(-14, 6, by = 4),
  lat = seq(57, 69, by = 3)
)
grd <- dist2land(grd, binary = TRUE, dist.col = "ocean",
                 verbose = FALSE)
grd <- grd[grd$ocean, ]

grd$u <- 0.35 + 0.25 * cos((grd$lat - 58) / 4) # eastward component
grd$v <- 0.20 * sin((grd$lon + 5) / 6)         # northward component

# Display scale only: a 1 m/s vector is drawn as 50 hours of drift.
arrow_hours <- 50
km_per_ms <- arrow_hours * 3.6
grd$lon_end <- grd$lon +
  (grd$u * km_per_ms) / (111.32 * cos(grd$lat * pi / 180))
grd$lat_end <- grd$lat + (grd$v * km_per_ms) / 110.57
grd <- dist2land(grd, lon = "lon_end", lat = "lat_end",
                 binary = TRUE, dist.col = "end_ocean",
                 verbose = FALSE)
grd <- grd[grd$end_ocean, ]

start <- transform_coord(grd, lon = "lon", lat = "lat",
                         proj.out = 3995, bind = TRUE)
end <- transform_coord(grd, lon = "lon_end", lat = "lat_end",
                       proj.out = 3995,
                       new.names = c("lon_end.proj", "lat_end.proj"))
grd <- cbind(start, end)

basemap(limits = c(-20, 30, 50, 70),
        bathy.style = "rbg", land.col = "grey86", legends = FALSE) +
  geom_segment(
    data = grd,
    aes(x = lon.proj, y = lat.proj,
        xend = lon_end.proj, yend = lat_end.proj),
    arrow = arrow(length = unit(0.12, "cm"), type = "open"),
    colour = "#084d9a",
    linewidth = 0.45
  )
```

On a decimal-degree map, the same code works without the
`transform_coord()` calls. Use decimal-degree maps only when the basemap
itself is in EPSG:4326; many high-latitude lon/lat limits automatically
select a polar projection.

# Pie charts

Pie charts can be drawn as ordinary polygons. This avoids an additional
plotting dependency and makes the coordinate units explicit: after
projection, the pie radius is in metres.

```{r pie-helper}
pie_polygons <- function(data, cols, x, y, r, n = 60) {
  out <- vector("list", nrow(data) * length(cols))
  k <- 1

  for (i in seq_len(nrow(data))) {
    values <- as.numeric(data[i, cols])
    values[is.na(values)] <- 0
    prop <- values / sum(values)
    starts <- c(0, cumsum(prop)[-length(prop)]) * 2 * pi
    ends <- cumsum(prop) * 2 * pi

    for (j in seq_along(cols)) {
      theta <- seq(starts[j], ends[j],
                   length.out = max(2, ceiling(n * prop[j])))
      out[[k]] <- data.frame(
        id = data$id[i],
        slice = cols[j],
        x = c(data[[x]][i], data[[x]][i] + cos(theta) * data[[r]][i],
              data[[x]][i]),
        y = c(data[[y]][i], data[[y]][i] + sin(theta) * data[[r]][i],
              data[[y]][i]),
        part = paste(data$id[i], cols[j], sep = "_")
      )
      k <- k + 1
    }
  }

  do.call(rbind, out)
}
```

```{r pie-projected, fig.height = 4.8}
pies <- data.frame(
  id = c("A", "B", "C", "D"),
  lon = c(-8, 0, 4, 9),
  lat = c(56.5, 60, 63.5, 67.5),
  cod = c(45, 20, 60, 25),
  haddock = c(30, 50, 15, 35),
  saithe = c(25, 30, 25, 40)
)
slices <- c("cod", "haddock", "saithe")

pies <- dist2land(pies, binary = TRUE, dist.col = "ocean",
                  verbose = FALSE)
stopifnot(all(pies$ocean))
pies <- transform_coord(pies, proj.out = 3995, bind = TRUE)
pies$r <- 90000 # metres
pie_df <- pie_polygons(pies, slices,
                       x = "lon.proj", y = "lat.proj", r = "r")

basemap(limits = c(-12, 16, 54, 70),
        bathy.style = "rbg", land.col = "grey86", legends = FALSE) +
  ggnewscale::new_scale_fill() +
  geom_polygon(
    data = pie_df,
    aes(x = x, y = y, group = part, fill = slice),
    colour = "white",
    linewidth = 0.15,
    alpha = 0.95
  ) +
  scale_fill_brewer(name = "Species", palette = "Set2")
```

If the map uses bathymetry, both bathymetry and pie slices use `fill`.
Add `ggnewscale::new_scale_fill()` between the basemap and the pie layer
when you need independent fill scales.

# Manual depth bins

The bathymetry layer uses a normal ggplot2 fill scale. Add another fill
scale after `basemap()` to replace the default palette and breaks.

```{r manual-continuous-bins, fig.height = 4.5, eval = file.exists(file.path(path.expand(getOption("ggOceanMaps.datapath", "")), "dd_rbathy_cont.rda"))}
basemap(c(11, 16, 67.3, 68.6),
        grid.col = NA, bathy.style = "rcb", downsample = 5) +
  scale_fill_stepsn(
    name = "Depth (m)",
    breaks = c(0, 50, 100, 200, 500, 1000),
    limits = c(0, NA),
    trans = "sqrt",
    colours = colorRampPalette(
      c("#F7FBFF", "#DEEBF7", "#9ECAE1", "#4292C6", "#08306B")
    )(5),
    na.value = "white"
  )
```

For the shipped binned raster, replace the discrete scale instead. Name
the vector to make the mapping independent of the number of bins present
in the plotted area.

```{r manual-discrete-bins, fig.height = 4.5}
depth_cols <- c(
  "0-50" = "#F7FBFF",
  "50-300" = "#C6DBEF",
  "300-500" = "#6BAED6",
  "500-1000" = "#2171B5",
  "1000-1500" = "#08306B",
  "1500-2000" = "#041836",
  "2000-4000" = "#02101F",
  "4000-6000" = "#000000",
  "6000-10000" = "#000000",
  "10000-Inf" = "#000000"
)

basemap(c(11, 16, 67.3, 68.6), grid.col = NA, bathy.style = "rbb") +
  scale_fill_manual(
    name = "Depth (m)",
    values = depth_cols,
    na.value = "white",
    drop = FALSE
  )
```

# Custom rasters with `raster_bathymetry()`

Use `raster_bathymetry()` when you want to preprocess a GEBCO, ETOPO,
IBCAO, or other raster before plotting. The example uses a small
synthetic `stars` raster, but the same call accepts a NetCDF or GeoTIFF
file path.

```{r raster-bathymetry-example, fig.height = 4.5}
library(stars)

lon <- seq(-5, 10, length.out = 80)
lat <- seq(50, 60, length.out = 60)
depth <- outer(
  lon, lat,
  function(x, y) 1200 + 600 * sin((x + 5) / 3) + 700 * cos((y - 50) / 2)
)
depth[depth < 0] <- 0

synthetic_bathy <- st_as_stars(
  list(depth = -depth),
  dimensions = st_dimensions(x = lon, y = lat)
)
st_crs(synthetic_bathy) <- 4326

rb <- raster_bathymetry(
  synthetic_bathy,
  depths = NULL,
  boundary = c(-5, 10, 50, 60),
  verbose = FALSE
)

basemap(
  limits = c(-5, 10, 50, 60),
  shapefiles = list(land = dd_land, glacier = NULL, bathy = rb),
  bathy.style = "rcb",
  grid.col = NA
) +
  scale_fill_gradientn(
    name = "Depth (m)",
    colours = colorRampPalette(
      c("#F7FBFF", "#9ECAE1", "#2171B5", "#08306B")
    )(8),
    limits = c(0, NA),
    na.value = "white"
  )
```

Set `depths` to a numeric vector to produce a binned raster for later
vectorization with `vector_bathymetry()`:

```{r raster-bathymetry-binned}
rb_binned <- raster_bathymetry(
  synthetic_bathy,
  depths = c(50, 200, 500, 1000, 2000),
  boundary = c(-5, 10, 50, 60),
  verbose = FALSE
)

rb_binned$depth.invervals
```
