---
title: "Adding graphical elements"
author: "Mikko Vihtakari (Institute of Marine Research)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Adding graphical elements}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
```

A `basemap()` is an ordinary `ggplot2` object: anything you can add to a
`ggplot()` with `+` works on a basemap too. This article collects two
graphical elements that come up often in oceanography but need a little care on
projected maps:

1. **Ocean-current arrows** drawn over a map.
2. **Pie charts (scatter pies)** placed at geographic positions.

The recurring theme is **coordinates**. On a decimal-degree map you plot in
longitude/latitude directly. On a *projected* map (polar regions, custom
`crs`) the axes are in metres, so geographic coordinates must be transformed
with `transform_coord()` first. Every recipe below shows both cases.

```{r setup}
library(ggOceanMaps)
library(ggplot2)
```

# Ocean-current arrows

There are two common situations: you have a **velocity field** (`u`/`v`
components on a grid) and want a quiver plot, or you have **hand-drawn
schematic arrows** (the "Figure 1" style arrows that summarise the general
circulation of a region).

## A velocity field as a quiver plot

The pattern is: thin the grid to a drawable density, scale the arrow length,
and draw with `geom_segment()`. The example uses a small synthetic field so it
runs without any download.

```{r quiver-dd, fig.height = 4.5}
set.seed(1)
grd <- expand.grid(lon = seq(-15, 25, by = 4),
                   lat = seq(52, 68, by = 2))
grd$u <- 0.6 * cos(grd$lat / 8)          # eastward component (m/s)
grd$v <- 0.3 * sin(grd$lon / 6)          # northward component (m/s)

scale <- 2                                # degrees drawn per m/s -- tune to taste
grd$lon_end <- grd$lon + grd$u * scale
grd$lat_end <- grd$lat + grd$v * scale

basemap(limits = c(-20, 30, 50, 70)) +
  geom_segment(
    data = grd,
    aes(x = lon, y = lat, xend = lon_end, yend = lat_end),
    arrow = arrow(length = unit(0.15, "cm")),
    colour = "steelblue", linewidth = 0.4
  )
```

On a **projected map** the axes are in metres, so transform the start points
and add the velocity in projected units:

```{r quiver-proj, eval = FALSE}
# transform_coord adds lon.proj / lat.proj in the map's CRS (metres)
grd <- transform_coord(grd, proj.out = 3995, bind = TRUE)

scale_m <- 1e5                            # metres drawn per m/s
grd$lon_end_proj <- grd$lon.proj + grd$u * scale_m
grd$lat_end_proj <- grd$lat.proj + grd$v * scale_m

basemap(60) +
  geom_segment(
    data = grd,
    aes(x = lon.proj, y = lat.proj, xend = lon_end_proj, yend = lat_end_proj),
    arrow = arrow(length = unit(0.15, "cm")),
    colour = "steelblue", linewidth = 0.4
  )
```

If your field lives in a NetCDF, read it with `stars::read_stars()`,
`as.data.frame()` it, rename the coordinate/velocity columns, and thin with
`df[seq(1, nrow(df), by = n), ]` before the steps above.

## Schematic "Figure 1" current arrows

For publication-style arrows summarising the circulation of a region, the
[Barents-Sea-currents](https://github.com/MikkoVihtakari/Barents-Sea-currents)
repository distributes ready-made, editable arrows for the Barents and
Greenland Seas. The arrows are stored as a small set of nodes (decimal
degrees) grouped per arrow, with a `type` column (`"Atlantic"` / `"Arctic"`).

The nodes are sparse on purpose (easy to edit); run an X-spline through each
arrow to smooth it before plotting. The recipe below reproduces the classic
two-colour current map on a ggOceanMaps basemap:

```{r currents-barents, eval = FALSE}
# 1. Read the editable node table (decimal degrees)
cur <- read.csv(paste0(
  "https://raw.githubusercontent.com/MikkoVihtakari/",
  "Barents-Sea-currents/master/tabular/barents_currents.csv"
))

# 2. Transform the nodes to the map projection (Arctic stereographic, metres)
cur <- transform_coord(cur, lon = "long", lat = "lat",
                       proj.out = 3995, bind = TRUE)

# 3. Smooth each arrow with an X-spline through its nodes.
#    xspline() needs an open graphics device even with draw = FALSE, so
#    open a null device for the calculation.
smooth_arrow <- function(d) {
  grDevices::pdf(NULL); on.exit(grDevices::dev.off()); graphics::plot.new()
  xs <- graphics::xspline(d$lon.proj, d$lat.proj, shape = -0.6, draw = FALSE)
  data.frame(lon.proj = xs$x, lat.proj = xs$y,
             group = unique(d$group), type = unique(d$type))
}
cur_smooth <- do.call(rbind, lapply(split(cur, cur$group), smooth_arrow))

# 4. Plot, mapping arrow colour to current type
basemap(limits = c(0, 50, 68, 83), shapefiles = "Arctic") +
  geom_path(
    data = cur_smooth,
    aes(x = lon.proj, y = lat.proj, group = group, colour = type),
    linewidth = 0.6,
    arrow = arrow(type = "open", angle = 15, ends = "last",
                  length = unit(0.3, "lines"))
  ) +
  scale_colour_manual(
    name = "Current type",
    values = c("Arctic" = "blue", "Atlantic" = "red")
  )
```

Key points:

- The CSV is in **decimal degrees**; `transform_coord(..., proj.out = 3995)`
  moves it to the Arctic-stereographic CRS that `basemap()` uses for this
  region. On a decimal-degree map skip the transform and map `aes(x = long,
  y = lat)` directly.
- `graphics::xspline(..., shape = -0.6, draw = FALSE)` returns smoothed
  coordinates without drawing anything — exactly what we feed to
  `geom_path()`.
- Smooth in the **projected** coordinates (after `transform_coord()`), not in
  degrees, so the curvature matches the map.
- If you cite these arrows, see the citation in the
  [source repository](https://github.com/MikkoVihtakari/Barents-Sea-currents#citations).

# Pie charts on maps (scatter pies)

A frequent request is a small pie chart at each station summarising a
composition (species proportions, water-mass fractions, …). The cleanest way
to do this on a `ggplot2` map is the
[**scatterpie**](https://cran.r-project.org/package=scatterpie) package, whose
`geom_scatterpie()` draws one pie per row of a data frame from a set of value
columns.

```{r, eval = FALSE}
install.packages("scatterpie")
```

Your data frame needs a position (`x`, `y`), an optional radius, and one
column per pie slice:

```{r pie-data, eval = FALSE}
pies <- data.frame(
  lon  = c(5, 15, 25),
  lat  = c(56, 62, 68),
  cod  = c(40, 10, 70),
  haddock = c(35, 30, 20),
  saithe  = c(25, 60, 10)
)
slices <- c("cod", "haddock", "saithe")
```

## On a decimal-degree map

Plot directly in longitude/latitude. `r` (the pie radius) is in **degrees**
here, because that is the unit of the map axes:

```{r pie-dd, eval = FALSE}
library(scatterpie)

basemap(limits = c(-5, 35, 52, 72)) +
  geom_scatterpie(
    data = pies,
    aes(x = lon, y = lat, r = 1.5),   # radius in degrees
    cols = slices,
    colour = NA, alpha = 0.9
  ) +
  scale_fill_brewer(palette = "Set2", name = "Species")
```

## On a projected map

Transform the pie centres first, and give `r` in the projected unit
(**metres** for the polar projections):

```{r pie-proj, eval = FALSE}
pies <- transform_coord(pies, proj.out = 3995, bind = TRUE)

basemap(60) +
  geom_scatterpie(
    data = pies,
    aes(x = lon.proj, y = lat.proj, r = 1.5e5),   # radius in metres
    cols = slices,
    colour = NA, alpha = 0.9
  ) +
  scale_fill_brewer(palette = "Set2", name = "Species")
```

## Scaling pies by a total and adding a size legend

Map the radius to a variable (e.g. total catch) and add a radius legend with
`geom_scatterpie_legend()`:

```{r pie-legend, eval = FALSE}
pies$total <- rowSums(pies[, slices])
pies$r <- 0.4 + 1.6 * pies$total / max(pies$total)   # degrees

basemap(limits = c(-5, 35, 52, 72)) +
  geom_scatterpie(
    data = pies,
    aes(x = lon, y = lat, r = r),
    cols = slices, colour = NA, alpha = 0.9
  ) +
  geom_scatterpie_legend(pies$r, x = -2, y = 70, n = 3) +
  scale_fill_brewer(palette = "Set2", name = "Species")
```

Tips:

- Keep `colour = NA` (no slice borders) for small pies; borders dominate at
  map scale.
- `geom_scatterpie()` adds a `fill` aesthetic, so it composes with the
  separate `fill` scale ggOceanMaps uses for bathymetry only if you add the
  pies *after* the bathymetry. If you need two independent fill scales (e.g.
  bathymetry **and** pie slices) use
  [**ggnewscale**](https://cran.r-project.org/package=ggnewscale)'s
  `new_scale_fill()` between the layers.
- Choose the radius in the same unit as the map axes: degrees on a
  decimal-degree map, metres on a projected map. A radius that looks right on
  one map will be invisible (or fill the plot) on the other.

# See also

- [Cookbook](cookbook.html) — short copy-pasteable recipes, including a
  compact version of the current-arrow recipe.
- [User manual](ggOceanMaps.html) — projections and `transform_coord()` in
  more depth.
