---
title: "Bathymetry in ggOceanMaps"
author: "Mikko Vihtakari (Institute of Marine Research)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Bathymetry in ggOceanMaps}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
```

```{r setup}
library(ggOceanMaps)
library(ggplot2)
```

ggOceanMaps can plot bathymetry from five different kinds of data source.
Each source has a sweet spot: some need nothing but the package itself,
others require an internet connection or a permanent download. This
vignette walks through all five in order of increasing setup cost.

# Quick decision guide

| Need | Recommended source | `bathy.style` |
|---|---|---|
| Anything, anywhere, no setup | Shipped low-res raster | `"rbb"` (default) |
| Higher detail, polar / global maps | ggOceanMapsLargeData | `"rcb"`, `"pb"`, `"cb"` |
| Your own GEBCO / ETOPO / IBCAO file | Local raster via `userpath` | `"rub"` |
| European waters, ~115 m | EMODnet WCS (live) | `"wemb"` |
| Anywhere on the globe, ~1.85 km | ETOPO1 WCS (live) | `"wceb"` |
| Custom contour polygons / land | `raster_bathymetry()` + `vector_bathymetry()` / `vector_land()` | `shapefiles = list(...)` |

Substitute `g` for the final `b` in any abbreviation to get the grey
variant (`rbb` → `rbg`, `wemb` → `wemg`, …). For the full alias list see
`?basemap`.

# 1. Shipped low-resolution raster (no setup)

The package bundles a coarse global bathymetry raster (`dd_rbathy`)
that's always available. This is the default — just toggle
`bathymetry = TRUE`:

```{r}
basemap(limits = c(-20, 30, 50, 70), bathymetry = TRUE)
```

Or set the style explicitly:

```{r}
basemap(c(11, 16, 67.3, 68.6), bathy.style = "rbb")
```

Greyscale variant:

```{r}
basemap(c(11, 16, 67.3, 68.6), bathy.style = "rbg")
```

The shipped raster is intentionally coarse so the package stays under
CRAN's size cap. It's fine for overview maps and quick exploratory
plots; for publication maps you'll usually want one of the higher-detail
options below.

# 2. ggOceanMapsLargeData (one-time download per region)

The companion repository
[ggOceanMapsLargeData](https://github.com/MikkoVihtakari/ggOceanMapsLargeData)
hosts higher-resolution rasters, polygon-contour bathymetries, and
contour lines for the supported regions (Decimal Degree, Arctic
Stereographic, Antarctic Stereographic, plus several pre-made regional
shapefile sets). `basemap()` downloads what it needs on first use and
caches it locally.

**One-time setup.** Pick a permanent directory (so files don't vanish
between R sessions) and add this line to `~/.Rprofile`:

```{r, eval = FALSE}
options(ggOceanMaps.datapath = "~/ggOceanMaps_data")
```

Then any high-res style will download into that directory and reuse it:

```{r, eval = FALSE}
# Continuous raster — recommended for most maps
basemap(c(11, 16, 67.3, 68.6), bathy.style = "rcb")

# downsample = n speeds up rendering at the cost of detail
basemap(c(11, 16, 67.3, 68.6), bathy.style = "rcb", downsample = 10)

# Greyscale variant
basemap(c(11, 16, 67.3, 68.6), bathy.style = "rcg")

# Polygon contours (pre-2.0 default)
basemap(c(11, 16, 67.3, 68.6), bathy.style = "pb")
basemap(c(11, 16, 67.3, 68.6), bathy.style = "pg")

# Plain contour lines
basemap(c(11, 16, 67.3, 68.6), bathy.style = "cb")
basemap(c(11, 16, 67.3, 68.6), bathy.style = "cg")
```

On the first call a menu asks you to confirm the download (~15–100 MB
depending on the region). Subsequent calls are instant.

# 3. Your own raster (GEBCO, ETOPO, IBCAO, …)

Sometimes you want a specific dataset — the latest GEBCO grid, an
ETOPO 2022 variant, a regional IBCAO compilation, or a custom-processed
file from your group. Download the file once, point ggOceanMaps at it
through `ggOceanMaps.userpath`, and use the `rub` / `rug` styles:

```{r, eval = FALSE}
# Set once (in .Rprofile for persistence):
options(ggOceanMaps.userpath = "path/to/your/bathymetry.nc")

# Then any basemap call uses your file:
basemap(c(11, 16, 67.3, 68.6), bathy.style = "rub")
basemap(c(11, 16, 67.3, 68.6), bathy.style = "rub", downsample = 10)
basemap(c(11, 16, 67.3, 68.6), bathy.style = "rug")
```

Any file format `stars::read_stars()` can open is accepted (NetCDF
`.nc`, GeoTIFF `.tif`, GMT `.grd`, …). The file is read fully every
time, so the user-raster route can be slower than a pre-processed
ggOceanMapsLargeData object for the same region. Use `downsample` to
trade resolution for speed.

`ggOceanMaps.userpath` can also be used by `get_depth()` to look up
point depths from your raster:

```{r, eval = FALSE}
dt <- data.frame(lon = seq(-20, 80, length.out = 41), lat = 50:90)
dt <- get_depth(dt, bathy.style = "ru")
qmap(dt, color = depth) + scale_color_viridis_c()
```

# 4. Live download from a Web Coverage Service (WCS)

For one-off maps or workflows where you don't want to manage local
files, ggOceanMaps can fetch bathymetry on demand from two OGC Web
Coverage Services. No pre-download needed; downloaded tiles are cached
under `getOption("ggOceanMaps.datapath")` so subsequent calls for the
same bbox are instant.

| Source | Style | Coverage | Resolution |
|---|---|---|---|
| **EMODnet** | `wcs_emodnet_blues` (`wemb`) | European waters | ~115 m |
| **ETOPO1 (NOAA NCEI)** | `wcs_etopo_blues` (`wceb`) | Global | ~1.85 km |

```{r, eval = FALSE}
# North Sea — high-resolution European waters from EMODnet
basemap(c(2, 3, 54, 55), bathy.style = "wemb")

# Hawaii — global coverage from ETOPO
basemap(c(-160, -154, 18, 23), bathy.style = "wceb")

# Indonesia / Java Trench — outside EMODnet, also ETOPO
basemap(c(110, 120, -20, 30), bathy.style = "wceb")
```

If you pick EMODnet for an area outside its European coverage, you get
a clear error pointing you to ETOPO:

```r
basemap(c(110, 120, -20, 30), bathy.style = "wemb")
#> Error: Bounding box (110.0° to 120.0° lon, -20.0° to 30.0° lat) lies
#> entirely outside the approximate coverage of EMODnet (≈-36° to 43° lon,
#> 15° to 90° lat).
#> For global bathymetry coverage, download GEBCO or ETOPO data locally and
#> use raster_bathymetry() + vector_bathymetry() instead, or wait for a
#> global WCS source to be added to ggOceanMaps.
```

(Replace `wemb` with `wceb` and the same call works.)

## Manual fetch with `wcs_bathymetry()`

The `bathy.style` route is the simplest, but for full control — passing
the raster into `vector_bathymetry()`, combining multiple regions,
sharing a cache with other tools — call `wcs_bathymetry()` directly:

```{r, eval = FALSE}
bathy <- wcs_bathymetry(c(2, 3, 54, 55), source = "emodnet")

basemap(c(2, 3, 54, 55),
        shapefiles = list(land = dd_land, glacier = NULL,
                          bathy = bathy$raster),
        bathymetry = TRUE)
```

## WCS caveats

- **Decimal-degree limits only.** Polar maps and projected-CRS limits
  are not supported (yet).
- **Per-source size caps.** EMODnet defaults to a 50 deg² maximum
  bounding box (it reads 8-byte doubles internally and a 4° tile already
  exceeds its read cap); ETOPO defaults to 2000 deg² because the
  underlying grid is much coarser. Override either with
  `max_area_deg2 = ...` in `wcs_bathymetry()`. Larger boxes are tiled
  and mosaicked automatically.
- **Citation.** EMODnet is CC-BY
  (<https://emodnet.ec.europa.eu/en/bathymetry>); ETOPO1 is Amante &
  Eakins 2009, NOAA NGDC
  (<https://www.ncei.noaa.gov/products/etopo-global-relief-model>). Cite
  the source when publishing figures.

# 5. Build your own shapefiles

If you need contour polygons, your own depth bins, or a matched
land + bathymetry pair, the `raster_bathymetry()` / `vector_bathymetry()`
/ `vector_land()` pipeline turns any raster source into reusable
shapefile objects:

```{r, eval = FALSE}
# 1. Process the raster — crop, sign-flip, optionally bin into contours
rb <- raster_bathymetry(
  "path/to/your/bathymetry.nc",
  depths = c(50, 200, 500, 1000, 2000, 4000),  # depth break points
  boundary = c(-5, 10, 50, 60)                  # crop to your region
)

# 2a. Vectorize bathymetry into depth-band polygons
vb <- vector_bathymetry(rb, drop.crumbs = 10)   # drop islands < 10 km²

# 2b. Vectorize land from the same raster
vl <- vector_land(rb, drop.crumbs = 10)

# 3. Plot — both layers plug into basemap()
basemap(c(-5, 10, 50, 60),
        shapefiles = list(land = vl, glacier = NULL, bathy = vb),
        bathymetry = TRUE)
```

`depths = NULL` skips the binning step and gives you a continuous
raster — useful for `geom_raster`-style fills without vectorization
overhead.

Save the processed objects so you don't pay the processing cost again:

```{r, eval = FALSE}
save(vb, vl, file = "my_region_bathy.rda")
```

# Picking a style at the bbox level

Switching `bathy.style` between maps in a multi-panel figure works as
expected; each panel uses its own source:

```{r, eval = FALSE}
library(patchwork)
p1 <- basemap(c(2, 3, 54, 55), bathy.style = "wemb") + ggtitle("EMODnet (115 m)")
p2 <- basemap(c(-160, -154, 18, 23), bathy.style = "wceb") + ggtitle("ETOPO (1.85 km)")
p1 + p2
```

# See also

- `?basemap` for the full `bathy.style` reference table.
- `?wcs_bathymetry` for the live download function.
- `?raster_bathymetry`, `?vector_bathymetry`, `?vector_land` for the
  build-your-own pipeline.
- `vignette("cookbook")` for short copy-pasteable recipes.
- The user manual: `vignette("ggOceanMaps")`.
