---
title: "ggOceanMaps cookbook"
author: "Mikko Vihtakari (Institute of Marine Research)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{ggOceanMaps cookbook}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
```

A collection of short, self-contained recipes for common ggOceanMaps tasks.
Each recipe is meant to be copy-pasteable. For background and the conceptual
walk-through, see the [user manual](ggOceanMaps.html).

```{r setup}
library(ggOceanMaps)
library(ggplot2)
```

# Choosing a region

## How do I plot a specific lon/lat box?

```{r}
basemap(limits = c(-20, 30, 50, 70))
```

`limits` is `c(xmin, xmax, ymin, ymax)`.

## How do I plot the Arctic / Antarctic?

A single value in `limits` means "polar projection bounded at this latitude":

```{r}
basemap(60)    # Arctic
```

```{r}
basemap(-60)   # Antarctic
```

## How do I plot a region that crosses the antimeridian?

```{r}
basemap(limits = c(160, -160, 60, 80), rotate = TRUE)
```

Without `rotate = TRUE` you'll get the whole world with a message; with it,
ggOceanMaps rotates the projection so the antimeridian sits in the middle.

## How do I fit the map to my data automatically?

```{r}
dt <- data.frame(lon = c(-150, 150), lat = c(60, 85))
basemap(data = dt, rotate = TRUE)
```

# Adding your own data

## How do I add points to a decimal-degree map?

```{r}
dt <- data.frame(lon = c(0, 10, 20), lat = c(55, 60, 65))
basemap(limits = c(-10, 30, 50, 70)) +
  geom_point(data = dt, aes(x = lon, y = lat), color = "red", size = 3)
```

## How do I add points to a *projected* map?

Projected maps (polar, custom CRS) need the coordinates transformed first:

```{r, eval = FALSE}
dt <- data.frame(lon = c(0, 10, 20), lat = c(70, 75, 80))
dt <- transform_coord(dt, bind = TRUE)   # adds lon.proj, lat.proj
basemap(60) +
  geom_point(data = dt, aes(x = lon.proj, y = lat.proj),
             color = "red", size = 3)
```

Or let `ggspatial` handle the transform:

```{r, eval = FALSE}
basemap(60) +
  ggspatial::geom_spatial_point(data = dt, aes(x = lon, y = lat),
                                crs = 4326, color = "red", size = 3)
```

## How do I quickly plot a data frame's coordinates?

```{r, eval = FALSE}
qmap(dt, color = I("red"))
```

`qmap()` picks limits and projection automatically.

# Bathymetry

## How do I add bathymetry without downloading anything?

```{r, eval = FALSE}
basemap(limits = c(-20, 30, 50, 70), bathymetry = TRUE)
```

This uses the binned-blues raster shipped with the package.

## How do I get higher-resolution bathymetry?

> *(Recipe stub — to be filled in alongside the WCS bathymetry work in
> phase 2: ggOceanMapsLargeData download flow, `options(ggOceanMaps.datapath)`,
> the `rcb` / `pb` / `cb` styles.)*

## How do I use my own bathymetry raster (e.g. GEBCO)?

> *(Recipe stub — `options(ggOceanMaps.userpath)`, `bathy.style = "rub"`.)*

## How do I fetch bathymetry on demand from EMODnet?

The `wcs_emodnet_blues` (abbrev. `wemb`) style downloads ~115 m
European-waters bathymetry from EMODnet's WCS endpoint on demand. No
pre-download needed; tiles are cached locally.

```{r, eval = FALSE}
basemap(limits = c(2, 3, 54, 55), bathy.style = "wcs_emodnet_blues")
```

For a manual fetch (e.g. to pass through `vector_bathymetry()` for a
custom shapefile workflow), use `wcs_bathymetry()` directly:

```{r, eval = FALSE}
bathy <- wcs_bathymetry(c(2, 3, 54, 55), source = "emodnet")
basemap(c(2, 3, 54, 55),
        shapefiles = list(land = dd_land, glacier = NULL,
                          bathy = bathy$raster),
        bathymetry = TRUE)
```

Requirements / caveats:

- **Decimal-degree limits only**; polar maps aren't supported by EMODnet.
- **European waters only**; EMODnet coverage is regional.
- **Bbox size is capped** at 10°² by default via `max_area_deg2`; raise
  it explicitly for larger requests.
- **Caching** uses `getOption("ggOceanMaps.datapath")` (or `tempdir()`
  as fallback). Set it permanently in `.Rprofile` for cross-session
  reuse.
- **Citation**: EMODnet bathymetry is CC-BY. Cite it when publishing
  figures — see <https://emodnet.ec.europa.eu/en/bathymetry>.

# Building your own shapefiles

## How do I make land + bathymetry from a downloaded raster?

> *(Recipe stub — `raster_bathymetry()` → `vector_bathymetry()` /
> `vector_land()` pipeline.)*

## How do I clip an existing shapefile to my region?

> *(Recipe stub — `clip_shapefile()`.)*

# Special cases

## How do I draw ocean current vectors over a map?

> *(Recipe stub — to be added with the Barents-Sea-currents example in
> phase 2.)*

## How do I add ICES or Norwegian fishery zones?

> *(Recipe stub — `ices_areas`, `fdir_main_areas`, `fdir_sub_areas`.)*

# Saving and exporting

## How do I save the map to PNG/PDF?

```{r, eval = FALSE}
p <- basemap(limits = c(-20, 30, 50, 70))
ggsave("europe.png", p, width = 6, height = 5, dpi = 300)
ggsave("europe.pdf", p, width = 6, height = 5)
```

`basemap()` returns a regular ggplot — anything that works on a ggplot works
here.
