---
title: "ggOceanMaps User Manual"
author: "Mikko Vihtakari (Institute of Marine Research)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  rmarkdown::html_vignette:
    toc_depth: 6
    fig_caption: true
vignette: >
  %\VignetteIndexEntry{ggOceanMaps User Manual}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE,
                      fig.pos = "H",
                      # fig.width = 9,
                      # fig.height = 5,
                      tab.pos = "H",
                      # collapse = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      comment = "#>"
)
```

The ggOceanMaps package for [R](https://www.r-project.org/) allows plotting data on bathymetric maps using [ggplot2](https://ggplot2.tidyverse.org/reference/). The package is designed for ocean sciences and greatly simplifies bathymetric map plotting anywhere around the globe. ggOceanMaps uses openly available geographic data. Citing the particular data sources is advised by the CC-BY licenses whenever maps from the package are published (see the [*Citations and data sources*](#citations-and-data-sources) section). Note that the package comes with absolutely no warranty and that maps generated by the package are meant for plotting scientific data only. The maps are coarse generalizations of third-party data and therefore inaccurate.

# Basic use

**ggOceanMaps** extends on [**ggplot2**](http://ggplot2.tidyverse.org/reference/). The package uses spatial ([**sf**](https://r-spatial.github.io/sf/)) shape- (e.g. vector) and ([**stars**](https://r-spatial.github.io/stars/)) raster files, [geospatial packages for R](https://cran.r-project.org/web/views/Spatial.html) to manipulate, and ggplot2 to plot these data. The vector and raster plotting is conducted internally in the `basemap` function, which uses [ggplot's sf object plotting capabilities](https://ggplot2.tidyverse.org/reference/ggsf.html).

The primary aim of ggOceanMaps is to make plotting oceanographic spatial data as simple as feasible, but also flexible for custom modifications. The "as simple as feasible" part will be covered in this section, while the "flexible for custom modifications" part is covered in the [Advanced use](#advanced-use) section. The basic use section of this tutorial assumes that the user knows how to use ggplot2. If you are not familiar with this package, you may read the [Data visualization](https://r4ds.had.co.nz/data-visualisation.html) section in [Hadley Wickham & Garrett Grolemund](https://r4ds.had.co.nz/). This tutorial does not describe functions in ggOceanMaps but rather focuses on how to use them. Make sure to refer to the [function documentation](https://mikkovihtakari.github.io/ggOceanMaps/reference/index.html) while reading the tutorial. 

```{r setup}
library(ggOceanMaps)
```

```{r, include = FALSE}
# Local build paths are read from environment variables so that no personal
# paths are committed. Set GGOCEANMAPS_DATAPATH (cached ggOceanMapsLargeData
# folder) and GGOCEANMAPS_USERPATH (your GEBCO/ETOPO grid) in ~/.Renviron to
# render the high-resolution / user-raster examples locally. Without them the
# examples that need downloads simply fetch to tempdir() during the build.
local({
  dp <- Sys.getenv("GGOCEANMAPS_DATAPATH")
  up <- Sys.getenv("GGOCEANMAPS_USERPATH")
  if (nzchar(dp)) options(ggOceanMaps.datapath = dp)
  if (nzchar(up)) options(ggOceanMaps.userpath = up)
})
```

## Limiting maps

To ensure simplicity, ggOceanMaps package attempts to use [decimal degree](https://en.wikipedia.org/wiki/Decimal_degrees) [coordinate system](https://en.wikipedia.org/wiki/Coordinate_system) as much as possible. This system represents coordinates on a sphere, while maps are plotted in two dimensions. Therefore, the underlying map data have to be projected using [different mathematical algorithms depending on the geographic location](#projections).

Maps can be limited (e.g. provide geographic location for a map) using three arguments: `limits`, `data`, and `shapefiles`. The limit type is automatically detected when supplied to the first argument (called `x`) in the `basemap()` and `qmap()` functions. 

### Limits

The simplest way of defining the geographic location is to use the `limits` argument with decimal degrees. The limits argument can be defined either as a numeric vector of length 1 or 4. Specifying the argument as a single integer between 30 and 88 or -88 and -30 plots a polar stereographic map for the Arctic or Antarctic, respectively.

```{r}
library(ggOceanMaps)
library(ggspatial) # for data plotting
basemap(limits = 60) # A synonym: basemap(60) 
```

Rectangular maps are plotted by specifying the `limits` argument as a numeric vector of length 4 where the first element defines the start longitude, the second element the end longitude, the third element the minimum latitude and the fourth element the maximum latitude of the bounding box:

```{r}
basemap(limits = c(-20, 20, 40, 59))
```

Limiting maps using decimal degrees is somewhat counter-intuitive because maps plotted for polar regions (>= 60 or <= -60 latitude) are actually projected to Arctic and Antarctic polar stereographic systems. Because decimal degrees are angular units running counter-clockwise, also the longitude limits have to be defined **counter-clockwise**. Projected maps with decimal degree `limits` will lead to expanded limits towards the poles when using Arctic and Antarctic Polar Stereographic projections because decimal degrees represent a sphere:

```{r, echo = FALSE}
start.lat <- data.frame(lon = seq(0, 360, 1), lat = 60)
end.lat <- data.frame(lon = seq(0, 360, 1), lat = 80)

start.lat.lon <- data.frame(lon = c(seq(120, 180, 1), seq(-180, -120, 1)), lat = 60)
end.lat.lon <- data.frame(lon = c(seq(120, 180, 1), seq(-180, -120, 1)), lat = 80)
lon.segments <- data.frame(lon = c(rep(-120, 2), rep(120, 2)), lat = rep(c(60, 80), 2))

real.lims <- auto_limits(expand.grid(lon = c(-120, 180, 120), lat = c(60, 60, 80)))
limits <- real.lims$projLimits

dt <- data.frame(lon = c(real.lims$projLimits[1:2]), lat = real.lims$projLimits[3:4])

labels.lat <- data.frame(lon = c(60, 60), lat = c(60, 80), label = c("limits[3]\n(Min. latitude)", "limits[4]\n(Max. latitude)"))
labels.lon <- data.frame(lon = c(120, -120), lat = c(rep(60, 2), rep(80, 2)), label = rep(c("limits[1]\n(Start longitude)", "limits[2]\n(End longitude)")))

p1 <- ggplot() +
  ggspatial::geom_spatial_path(
    data = start.lat, 
    aes(x = lon, y = lat), crs = 4326, linetype = 2) +
  ggspatial::geom_spatial_path(
    data = end.lat, 
    aes(x = lon, y = lat), crs = 4326, linetype = 2) +
  ggspatial::geom_spatial_path(
    data = start.lat.lon, 
    aes(x = lon, y = lat), crs = 4326, color = "red") +
  ggspatial::geom_spatial_path(
    data = end.lat.lon, 
    aes(x = lon, y = lat), crs = 4326, color = "red") +
  ggspatial::geom_spatial_path(
    data = lon.segments, 
    aes(x = lon, y = lat, group = lon), crs = 4326, color = "red") +
  ggspatial::layer_spatial(data = real.lims$projBound, 
                           fill = NA, color = "red", size = 2) +
  scale_x_continuous("Longitude (decimal degrees)", breaks = seq(0, 360, 30)) +
  geom_label(data = data.frame(lon = 0, lat = 0, label = "The pole"), 
             aes(x = lon, y = lat, label = label), size = FS(8)) +
  ggspatial::geom_spatial_label(
    data = labels.lat, 
    aes(x = lon, y = lat, label = label), crs = 4326, size = FS(8)) +
  ggspatial::geom_spatial_label(
    data = labels.lon, 
    aes(x = lon, y = lat, label = label), 
    crs = 4326, color = "red", size = FS(8)) +
  ylab("Longitude (decimal degrees)") + 
  coord_sf(default_crs = NULL, crs = 3995, label_axes = "EEEE") +
  
  theme_map(grid.col = "grey70", grid.size = 0.1, base_size = 8)

labels.proj <-
  data.frame(
    lon = c(limits[1], limits[2], mean(limits[1:2]), mean(limits[1:2])),
    lat = c(mean(limits[3:4]), mean(limits[3:4]), limits[3], limits[4]),
    label = c("limits[1]\n(Min. longitude)", "limits[2]\n(Max. longitude)",
              "limits[3]\n(Min. latitude)", "limits[4]\n(Max. latitude)")
  )

p2 <- ggplot() +
  geom_vline(xintercept = real.lims$projLimits[1:2], linetype = 2) +
  geom_hline(yintercept = real.lims$projLimits[3:4], linetype = 2) +
  ggspatial::layer_spatial(data = real.lims$projBound, fill = NA, 
                           color = "red", size = 2) +
  geom_label(data = labels.proj, 
             aes(x = lon, y = lat, label = label), color = "red", size = FS(8)) +
  geom_label(data = data.frame(lon = 0, lat = 0, label = "The pole"), 
             aes(x = lon, y = lat, label = label), size = FS(8)) +
  geom_spatial_path(data = start.lat, 
                    aes(x = lon, y = lat), crs = 4326, color = NA) +
  labs(y = "Latitude (meters from the pole)", x = "Longitude (meters from the pole)") +
  coord_sf(default_crs = NULL, crs = 3995, datum = 3995) +
  coord_sf(crs = 3995, datum = 3995) +
  theme_map(grid.col = "grey70", grid.size = 0.1, base_size = 8) + 
  theme(plot.margin = margin(t = 20, unit = "pt"))

cowplot::plot_grid(p1, p2, labels = "AUTO", align = "hv")
```

The figure above: Limiting rectangular basemaps is done by placing four coordinates to the limit argument. A) If the limits are in decimal degrees, the longitude limits (`[1:2]`) specify the start and end segments of corresponding angular lines that should reside inside the map area. The longitude limits are defined **counter-clockwise**. The latitude limits `[3:4]` define the parallels that should reside inside the limited region given the longitude segments. Note that the resulting limited region (polygon with thick red borders) becomes wider than the polygon defined by the coordinates (thin red borders). The example limits are `c(120, -120, 60, 80)`. B) If the limits are given as projected coordinates or as decimal degrees for maps with |latitude| < 60, limits elements represent lines encompassing the map area in cartesian space. The example limits are the limits from A) projected to the Arctic stereographic (crs = 3995). 

The example above as a map. Note how the latitude limits give a much larger map than one would expect from cartesian coordinates because the 60 $^\circ$N parallel is within the map area between 120 $^\circ$E and 120 $^\circ$W meridians. 

```{r, fig.height=4}
dt <- data.frame(lon = c(120, 120, -120, -120), lat = c(60, 80, 80, 60))

basemap(limits = c(120, -120, 60, 80)) +
  ggspatial::geom_spatial_polygon(
    data = dt, 
    aes(x = lon, y = lat), fill = NA, color = "red")
```

Exact control of map limits can be difficult using decimal degree limits in polar regions. The `limits` argument also allows specifying the limits in the underlying projected coordinate units. First, we will need to find out how these units look like:

```{r}
basemap(limits = 60, projection.grid = TRUE, grid.col = "red")
```

The `projection.grid` argument plots a grid using the projected actual map coordinates instead of decimal degrees. The grid helps in defining the `limits` using projected coordinates giving better control over the map limits than decimal degree coordinates. The automatic shapefile definition algorithm does not work for projected coordinates. Therefore, if the limits are not given as decimal degrees (any longitude outside the range [-180, 180] or latitude [-90, 90]), the function asks to specify `shapefiles`. The `shapefiles` can be defined by partially matching the names of the pre-made shapefiles in `shapefile_list` (e.g. "Ar" would be enough for "ArcticStereographic"):

```{r}
basemap(limits = c(-2e6, 1e6, 0, 3e6), shapefiles = "Arctic") 
```

### Data limits

The limits of a map can also be defined by inputting a data frame to the `data` argument. The limits are automatically defined allowing the user to quickly find limits for a desired spatial dataset:

```{r}
dt <- expand.grid(lon = c(160, 180, -160), lat = c(60, 70, 80))

basemap(data = dt) + # a synonym: basemap(dt)
  ggspatial::geom_spatial_point(data = dt, aes(x = lon, y = lat), color = "red")
```

The function makes the map such that the outermost data points barely fit into the mapped region. The space between the map borders and data points can be adjusted using the `expand.factor` argument:

```{r fig-expand-factor, fig.cap = "Figure above: The expand.factor argument can be used to expand (A) and reduce (B) map region in relation to data."}
cowplot::plot_grid(
  basemap(dt, expand.factor = 1.1) +
    ggspatial::geom_spatial_point(data = dt, aes(x = lon, y = lat), color = "red") +
    theme(axis.title = element_blank()),
  basemap(dt, expand.factor = 0.9) +
    ggspatial::geom_spatial_point(data = dt, aes(x = lon, y = lat), color = "red") +
    theme(axis.title = element_blank()),
  labels = "AUTO"
)
```

See the [Adding data to maps](#adding-data-to-maps) section for more information. The `basemap()` function automatically detects columns containing longitude and latitude information when the `data` argument is used. The automatic detection algorithm is not very advanced and it is recommended to use intuitive column names for longitude (such as "lon", "long", or "longitude") and latitude ("lat", "latitude") columns. The coordinate data have to be given as decimal degrees for the `data` argument to function. 

### Shape files

The ggOceanMaps package contains a set of [premade shapefiles]. Only the decimal degree land shapes are shipped with the package while others are downloaded as needed. See the [front page for instructions how to setup the automatic download folder](https://mikkovihtakari.github.io/ggOceanMaps/index.html#data-path) before you start downloading shape files. Any character supplied to the `x` argument in the `basemap()` function will automatically be understood as `shapefiles` argument. The maps are limited showing the entire land shape. Use the `limits` argument to further limit the shape file. 

```{r}
basemap("Arctic")
```

## Bathymetry

The simplest way to add bathymetry is `bathymetry = TRUE`, which uses
the low-resolution raster shipped with the package:

```{r}
basemap(limits = c(-125, -110, 25, 35), bathymetry = TRUE)
```

For higher detail, set `bathy.style` to one of the alternative styles.
The default is `"raster_binned_blues"` (`"rbb"`); switching to
`"raster_continuous_blues"` (`"rcb"`) gives you a higher-resolution
continuous raster (first call downloads it from
[ggOceanMapsLargeData](https://github.com/MikkoVihtakari/ggOceanMapsLargeData)).

```{r, eval = FALSE}
basemap(limits = c(-125, -110, 25, 35), bathy.style = "rcb")
```

ggOceanMaps supports five different bathymetry data sources:

1. **Shipped low-resolution raster** — `"rbb"` / `"rbg"`. Default. No
   setup, works offline (downsampled from ETOPO 2022).
2. **ggOceanMapsLargeData** higher-resolution data — a continuous raster
   (`"rcb"`), filled polygon contours (`"pb"`), or contour lines
   (`"cb"`). One-time download per region.
3. **Live download from a Web Coverage Service** — `"wceb"` (ETOPO1,
   ~1.85 km global) or `"wemb"` (EMODnet, ~115 m European waters). Tiles
   are cached locally after the first fetch.
4. **Your own raster** (GEBCO, ETOPO, IBCAO, …) via
   `options(ggOceanMaps.userpath = "...")` — `"rub"` / `"rug"`.
5. **Build-your-own** with `raster_bathymetry()` →
   `vector_bathymetry()` / `vector_land()`.

The `bathy.style` string follows the pattern `geometry_palette`. Add
`_grays` (or change the final `b` → `g` in the abbreviation) for the
greyscale variant of any style. The full reference table — every style,
every abbreviation, and what each one needs — lives in `?basemap`. Each
source is walked through with examples, and the data sources to cite are
listed, in the dedicated [Bathymetry vignette](bathymetry.html).

The *default `bathy.style` can be changed by setting the `ggOceanMaps.bathy.style` option*. `options(ggOceanMaps.bathy.style = "poly_blues")` would make the style similar to older pre-2.0 versions of ggOceanMaps.

## Glaciers

Since 2.0, glaciers require a download. It is a good idea to use the polar stereographic datasets for this purpose:

```{r}
basemap(limits = -60, glaciers = TRUE, shapefiles = "Antarctic")
```

## Adding data to maps {#adding-data-to-maps}

The `basemap(...)` function works almost similarly to the `ggplot(...)` function as a base for adding further layers to the plot. The difference between the `basemap()` and the `ggplot()` is that the `basemap()` plot already contains multiple ggplot layers. All layers except bathymetry have no other `aes` mapping than `x`, `y` and `group`. Bathymetry is mapped to `fill` or `color` color in addition. This means that when you add ggplot layers, you need to specify the `data` argument explicitly as shown below. Another difference is that basemaps are plotted using projected coordinates. The [ggspatial](https://CRAN.R-project.org/package=ggspatial) and ggplot's [`geom_sf`](https://ggplot2.tidyverse.org/reference/ggsf.html) functions convert the coordinates automatically to the projected coordinates:

```{r}
dt <- data.frame(lon = c(seq(-180, 0, 30), seq(30, 180, 30)), lat = -70)

basemap(limits = -60, glaciers = TRUE, shapefiles = "Antarctic") + 
  ggspatial::geom_spatial_point(data = dt, aes(x = lon, y = lat), color = "red")
```

The ggplot functions can also be used, but the coordinates need to be transformed to the basemap projection first using the `transform_coord` function:

```{r}
basemap(limits = -60, glaciers = TRUE, shapefiles = "Antarctic") + 
  geom_point(data = transform_coord(dt), aes(x = lon, y = lat), color = "red")
```

Note that the maps plotted in temperate and tropical regions are not projected. Consequently, decimal degrees work for such maps directly:

```{r}
dt <- data.frame(lon = c(-100, -80, -60), 
                 lat = c(10, 25, 40), 
                 var = c("a", "a", "b")
)

basemap(data = dt) + 
  geom_point(data = dt, aes(x = lon, y = lat), color = "red")
```

The `transform_coord` function detects the projection automatically, given that the map is limited using a similar range of coordinates. Therefore you can use the `transform_coord` as demonstrated above whenever using standard ggplot layers.

```{r}
transform_coord(data.frame(lon = -80, lat = 25), bind = TRUE)
```

## Rotating maps

The stereographic maps can be rotated to point towards north using the `rotate` argument:

```{r}
basemap(limits = c(-160, -80, 60, 85), rotate = TRUE)
```

Note that rotation changes the underlying CRS and data need to be added using `ggspatial::geom_spatial_*`, `ggplot2::geom_sf()` or `transform_coord(rotate = TRUE)` functions.

## Quick map

The `qmap` function is designed as a shortcut to quickly take a look at a spatial dataset. This function automatically detects the type of data fed into the function and plots a map using appropriate geometries, limits, and projection. You can use the `expand.factor` argument to adjust the automatic zoom into `data`. 

```{r}
dt <- data.frame(lon = c(-100, -80, -60), 
                 lat = c(10, 25, 40), 
                 var = c("a", "a", "b")
)

qmap(dt, color = I("red")) # set color
qmap(dt, color = var, expand.factor = 1.3) # map color, zoom out
```

# Advanced use

This section focuses on flexibility and user modifications. It is assumed that advanced users understand the basics of geographic information systems (GIS) and how to use these systems in R (e.g. see the [Making Maps with R](https://geocompr.robinlovelace.net/adv-map.html) chapter in [Lovelace et al. 2020](https://geocompr.robinlovelace.net/index.html)). 

## Projections {#projections}

The `basemap` function uses the `limits` argument to automatically detect the required projection for a map (or the `data` argument to calculate `limits`). The algorithms deciding which projection to use are defined in `define_shapefiles` and `shapefile_list` functions. The `basemap` function uses decimal degree land shapes as default and transforms them to polar stereographic projections based on following rules (given as [EPSG codes](https://epsg.io/)):

- **4326** WGS 84 / World Geodetic System 1984, used in GPS. Called "DecimalDegree". For min latitude (`limits[3]`) < 30 or > -30, max latitude (`limits[4]`) < 60 or > -60, and single integer latitudes < 30 and > -30.
- **3995** WGS 84 / Arctic Polar Stereographic. Called "ArcticStereographic". For max latitude (`limits[4]`) >= 60 (if min latitude (`limits[3]`) >= 30), and single integer latitudes >= 30 and <= 89.
- **3031** WGS 84 / Antarctic Polar Stereographic. Called "AntarcticStereographic". For max latitude (`limits[4]`) <= -60 (if min latitude (`limits[3]`) <= -30), and single integer latitudes <= -30 and >= -89.

Further, there are higher resolution data that can be downloaded when needed. They use a projection which suits that region. Here are all `shapefiles` and their projections (CRS):

```{r, echo = FALSE}
knitr::kable(shapefile_list("all")[c("name", "crs")])
```

Since 2.0, it is possible to override the default CRS used by `basemap()` using the `crs` argument:

```{r}
basemap(limits = c(0, 15, 55, 65), crs = 32631)
```

## Appearance

### Customizing bathymetry scales

The `bathy.style = "*_binned_*"` bathymetry polygons are mapped to `geom_fill_discrete` and can be modified using standard ggplot syntax:

```{r}
basemap(limits = c(-140, -105, 20, 40), bathymetry = TRUE) + 
  scale_fill_viridis_d("Water depth (m)")
```

The `bathy.style = "*_continuous_*"` bathymetry polygons are mapped to `geom_fill_continuous`:

```{r, eval = FALSE}
basemap(limits = c(-140, -105, 20, 40), bathy.style = "rcb") +
  scale_fill_viridis_c("Water depth (m)")
```

The `bathy.style = "contour_*"` bathymetry lines are mapped to `geom_color_discrete`:

```{r}
basemap(limits = c(0, 60, 68, 80), bathymetry = TRUE, bathy.style = "contour_blues") + 
  scale_color_hue()
```

## Graphical parameters

The `basemap` function uses graphical parameters that (very objectively) happen to please the eye of the author and have worked in the applications needed by the author. The default parameters may suddenly change without warning. You may want to modify the appearances of a `basemap` to your own liking. This can be done using the `*.col` (fill), `*.border.col` (line color) and `*.size` (line width) arguments:

```{r}
basemap(limits = c(-20, 30, 55, 70), glaciers = TRUE, 
        bathymetry = TRUE, bathy.style = "poly_greys",
        land.col = "#eeeac4", gla.col = "cadetblue", 
        land.border.col = NA, gla.border.col = NA,
        grid.size = 0.05)
```

Graticules (the grid lines) can be removed by setting the `grid.col` to `NA`. Axis labels can be manipulated using standard ggplot code:

```{r}
basemap(limits = c(124, 148, 31, 50), grid.col = NA) + 
  labs(x = NULL, y = "Only latitude for you, ...")
```

## Add scale bar and north arrow

Scale bar and north arrows can be added using the [**ggspatial** functions](https://paleolimbot.github.io/ggspatial/index.html):

```{r}
basemap(limits = c(-75, -45, 62, 78), rotate = TRUE) + 
  ggspatial::annotation_scale(location = "br") + 
  ggspatial::annotation_north_arrow(location = "tr", which_north = "true") 
```

Note that the north arrow in the example above points toward north where it is placed and that the direction of north varies as shown by the meridians. The scale bar is correct at 71 $\circ$N latitude as specified by the projection (crs = 3995) for Arctic stereographic maps. 

## Modifying basemap objects

The objects produced by the `basemap` function are standard ggplot objects with the difference that relevant information used in mapping is added to `attributes` of the object:

```{r}
p <- basemap(-60)
names(attributes(p))
```

Accessing the `attributes` allow custom modifications of maps produced by the `basemap` function. See the [Reordering layers](#reordering-layers) section as an example. 

## Reordering layers

Sometimes there is a need to move land, glacier, and grid layers on top of spatial data added on a `basemap`. This can be done using the `reorder_layers` function. This example uses Norwegian fishing regions (Hovedområder f.o.m. 2018), which can be downloaded from the [Norwegian Directorate of Fisheries data portal](https://www.fiskeridir.no/) (use "ESRI shapefile" option). The example works for any spatial polygons with crs information, however.

The Norwegian fishing regions are included as an example dataset in the **ggOceanMaps** package. You can add any data you want using the `sf::st_read` function.

```{r}
data(fdir_areas)

basemap(fdir_main_areas) + 
  ggspatial::annotation_spatial(fdir_main_areas, fill = NA)
```

The initial plot draws the polygons. Note how the polygon boundaries are partly on land. We want to eventually hide them under land. We also add region labels and color the polygons based on their area to demonstrate the capabilities of sf, ggplot2, ggspatial and ggOceanMaps: 

```{r}
labels <- suppressWarnings(sf::st_centroid(fdir_main_areas))
fdir_main_areas$area <- as.numeric(sf::st_area(fdir_main_areas))/1e9 # calculate area in 1000 km2

p <- basemap(fdir_main_areas) +
  geom_sf(data = fdir_main_areas, aes(fill = area)) +
  geom_sf_text(data = labels, aes(label = main_area), size = FS(8), fontface = 2) +
  scale_fill_distiller(name = "Area\n(1000 km2)",
                       palette = "Spectral", na.value = "white",
                       limits = c(0, 500), oob = scales::squish)

reorder_layers(p)
```

Ideally, the region labels should not go under land. This can be fixed by plotting the labels on top of the reordered ggplot object. To demonstrate how to reorder layers, we do this manually here:

```{r}
p <- reorder_layers(p)
tmp <- sapply(p$layers, function(k) !is.null(k$mapping$label)) # the layer with label mapping
p$layers <- c(p$layers[-which(tmp)], p$layers[which(tmp)])
p
```

## Contolling the plotting order of graticules

Graticules are plotted using [`ggplot2::coord_sf()`](https://ggplot2.tidyverse.org/reference/ggsf.html) and are regarded as the *panel* element in the [`ggplot2::theme()`](https://ggplot2.tidyverse.org/reference/theme.html). Note that also background color is included in the *panel* element (see the next section). The [`ggOceanMaps::theme_map()`](https://mikkovihtakari.github.io/ggOceanMaps/reference/theme_map.html) function, which is automatically executed together with every `basemap()` and `qmap()` command, controls the position of the graticules and they are set on top by default:

```{r}
basemap(limits = c(-20, 15, 50, 70), grid.col = "red", grid.size = 1)
```

To move the graticules under all layers, use the `panel.ontop` argument in `ggplot2::theme()`:

```{r}
basemap(limits = c(-20, 15, 50, 70), grid.col = "red", grid.size = 1) + 
  theme(panel.ontop = FALSE)
```

Sometimes there is a need to control the position of graticules in between layers. At the time of writing, ggplot does not offer a straightforward way to do this, but we can use the code from [`ggplot2::coord_sf()`](https://github.com/tidyverse/ggplot2/blob/7571122f6a9455ccce9f8ceb9a2196c0a70be646/R/coord-sf.R#L226) to make the graticules:

```{r}
p <- basemap(limits = c(-20, 15, 50, 70), grid.col = NA) + # without graticules 
  geom_sf(data = ices_areas, aes(fill = SubArea), show.legend = FALSE) 

# Make the graticules:
lims <- attributes(p)$limits 
graticule <- sf::st_graticule(
  c(lims[1], lims[3], lims[2], lims[4]), 
  crs = attributes(p)$crs,
  lon = seq(-180, 180, 15), lat = seq(-90, 90, 5))

# Plot
reorder_layers(p) +
  geom_sf(data = graticule, color = "grey70", size = LS(0.5)) + # graticules
  coord_sf(xlim = lims[1:2], ylim = lims[3:4], # redefine limits
           crs = attributes(p)$crs) +
  # labels on top of graticules
  geom_sf_label(
    data = suppressWarnings(sf::st_centroid(sf::st_make_valid(ices_areas))), 
    aes(label = Area_Full)
  )
```

An exception are the polar maps for which graticules are drawn by ggOceanMaps (because ggplot2 graticules for those maps were bugged some years ago when writing the function; this part requires a revision at some point). 

```{r}
X <- basemap_data(60)

basemap(60, grid.col = NA) +
  geom_sf(
    data = X$map.grid$lon.grid.lines, 
    color = "red", size = 1) +
  geom_sf(
    data = X$map.grid$lat.grid.lines, 
    color = "red", size = 1) +
  geom_sf(
    data = X$shapefiles$land, 
    fill = "grey60", color = "black", size = 0.1)
```

## Coloring the ocean (background)

Similarly, the [`ggplot2::theme()`](https://ggplot2.tidyverse.org/reference/theme.html) *panel* element allows [coloring the ocean](https://stackoverflow.com/questions/67615318/how-to-colour-the-ocean):

```{r}
basemap(expand.grid(lon = c(-129, -124), lat = c(49, 53)), grid.col = "red", grid.size = 0.5) + 
  theme(panel.background = element_rect(fill = "lightblue"),
        panel.ontop = FALSE) 
```

If you want the graticules on top, you'll need to define them manually because, both, the background and the grid, are contained in the same ggplot element called *panel*:

```{r}
p <- basemap(expand.grid(lon = c(-129, -124), lat = c(49, 53)), grid.col = NA) + 
  theme(panel.background = element_rect(fill = "lightblue"),
        panel.ontop = FALSE) 

# Make the graticules:
lims <- attributes(p)$limits 
graticule <- sf::st_graticule(
  c(lims[1], lims[3], lims[2], lims[4]), 
  crs = attributes(p)$crs,
  lon = attributes(p)$map.grid$lon.breaks, 
  lat = attributes(p)$map.grid$lat.breaks
)

# Plot
p + 
  geom_sf(data = graticule, color = "red", size = LS(1)) + # graticules
  coord_sf(xlim = lims[1:2], ylim = lims[3:4], # redefine limits
           crs = attributes(p)$crs) 
```

## Add a fill scale on top of bathymetry

```{r eval = FALSE}
basemap(c(-20, 15, 50, 70), bathymetry = TRUE) + 
  annotation_spatial(ices_areas, aes(fill = Area_Full))
#> Error in `f()`:
# ! Insufficient values in manual scale. 75 needed but only 9 provided.
# Run `rlang::last_error()` to see where the error occurred.
```

This error comes because ggplot does not allow two color scales with similar mapping in one plot. The issue can be evaded by using contour bathymetry:

```{r}
basemap(c(-20, 15, 50, 70), bathymetry = TRUE, 
        bathy.style = "contour_blues", legends = FALSE) + 
  annotation_spatial(ices_areas, aes(fill = Area_Full), alpha = 0.4) + 
  theme(legend.position = "none")
```

Or by using the [ggnewscale](https://github.com/eliocamp/ggnewscale) package to make the data `fill` mapping disconnected from that in `basemap`:

```{r}
basemap(limits = c(-20, 15, 50, 70), bathymetry = TRUE, 
        bathy.style = "poly_greys") +
  ggnewscale::new_scale_fill() +
  annotation_spatial(ices_areas, aes(fill = Area_Full), alpha = 0.4) + 
  theme(legend.position = "none")
```

## Add longitude and latitude labels to a polar map

Since adding longitude and latitude labels to polar maps requires some manual fiddling for each map, there is no premade solution for this. Instead, you can do this using the sf package:

```{r}
library(sf); library(dplyr)

p <- basemap(-60, shapefiles = "Antarctic")

p + 
  geom_text(
    data =
      attributes(p)$map.grid$lat.grid.lines %>% 
      sf::st_coordinates() %>% 
      as.data.frame() %>% 
      dplyr::filter(X == -180) %>%
      dplyr::add_row(X = -180, Y = -90, L1 = 3, L2 = 1) %>% 
      transform_coord(proj.out = attributes(p)$crs, bind = TRUE),
    aes(x = lon.proj, y = lat.proj, label = paste0(abs(Y), "\u00B0S"))
  ) +
  geom_text(
    data =
      attributes(p)$map.grid$lon.grid.lines %>% 
      sf::st_coordinates() %>% 
      as.data.frame() %>% 
      dplyr::filter(Y == -60) %>%
      dplyr::mutate(Y = -63) %>% 
      transform_coord(proj.out = attributes(p)$crs, bind = TRUE),
    aes(x = lon.proj, y = lat.proj, 
        label = ifelse(sign(X) == 1, paste0(abs(X), "\u00B0E"), 
                       paste0(abs(X), "\u00B0W"))
    )
  ) +
  theme(plot.margin = margin(0.25, 0.25, 0.25, 0.25, "cm"))
```

## Custom shapefiles

When the premade maps do not cover your region at the resolution you need, you
can pass your own land, glacier, and bathymetry layers to `basemap()` via the
`shapefiles` argument:

```{r, eval = FALSE}
basemap(
  limits = c(10, 53, 70, 80),
  shapefiles = list(land = my_land, glacier = my_glacier, bathy = my_bathy),
  bathymetry = TRUE, glaciers = TRUE
)
```

The list elements `land`, `glacier`, and `bathy` are all recognised; set
`glacier` and/or `bathy` to `NULL` if you do not need them. All three must
share one projection, which `basemap()` then uses for the map.

Building those layers — clipping an existing shapefile, turning a GEBCO/ETOPO
grid into matched land and depth-contour polygons with `raster_bathymetry()` →
`vector_bathymetry()` / `vector_land()`, or reading Geonorge depth data with
`geonorge_bathymetry()` — is covered step by step in the dedicated
[Customising shapefiles](customising-shapefiles.html) article. Useful raster
and vector data sources are listed in the
[Bathymetry](bathymetry.html) article.

# Known issues

The land and glacier shapes do not get dissolved correctly when projecting from decimal degrees:

```{r}
basemap(60, glaciers = TRUE)
```

A solution until the issue has been fixed is to use projected downloadable shape files:

```{r}
basemap(60, glaciers = TRUE, shapefiles = "Arctic")
```

# Citations and data sources

The data used by the package are not the property of the Institute of Marine Research nor the author of the package. It is, therefore, important that you cite the data sources used in a map you generate with the package. Please see [here](https://mikkovihtakari.github.io/ggOceanMaps/#citations-and-data-sources) for a list of data sources. 

Please cite the package whenever maps generated by the package are published. For up-to-date citation information, please use:

```{r}
citation("ggOceanMaps")
```
