---
title: "New features in ggOceanMaps version 3"
author: "Mikko Vihtakari (Institute of Marine Research)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{New features in ggOceanMaps version 3}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  eval = FALSE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
```

```{r setup, eval = TRUE}
library(ggOceanMaps)
```

# Bringing ggOceanMaps to the AI age

Version 3 marks a change in direction. From now on, ggOceanMaps is developed
with the help of AI coding assistants (Claude Code, GitHub Copilot), and a core
goal of the package is to **help AI agents make good maps on behalf of their
users**. Plotting oceanographic data should be a one-sentence request, whether
you type it yourself or ask an assistant to do it for you.

The most visible part of this shift is a new
[`AGENTS.md`](https://github.com/MikkoVihtakari/ggOceanMaps/blob/master/AGENTS.md)
file shipped with the package. It gives AI assistants concise, accurate
instructions for *using* ggOceanMaps — the projection rules, the `shapefiles`
contract, how `transform_coord()` works, and the common pitfalls — so they spend
less time guessing and more time helping. The documentation site was also
reorganised around this idea: a short [user manual](ggOceanMaps.html) that links
out to focused, in-depth articles.

Everything below is new or substantially improved in version 3.

# On-demand high-resolution bathymetry

The biggest functional addition is `wcs_bathymetry()`: high-resolution
bathymetry downloaded on demand from Web Coverage Services, with no manual file
handling. You select it through the `bathy.style` argument and ggOceanMaps
fetches, caches, and renders the right tiles for your map extent.

Two sources are available:

- **EMODnet** (~115 m, European waters) — `bathy.style = "wcs_emodnet_blues"`
  (`"wemb"`) or `"wcs_emodnet_grays"` (`"wemg"`).
- **ETOPO1 from NOAA NCEI** (~1.85 km, global) — `bathy.style =
  "wcs_etopo_blues"` (`"wceb"`) or `"wcs_etopo_grays"` (`"wceg"`). Use this when
  EMODnet has no coverage.

```{r}
# High-resolution EMODnet bathymetry around Tromsø, downloaded on the fly
basemap(c(18.4, 19.3, 69.5, 69.9), bathy.style = "wemb")
```

Bounding boxes outside a source's coverage fail cleanly with a pointer to the
right alternative, large boxes are tiled and mosaicked automatically, and the
downloaded tiles are cached under `getOption("ggOceanMaps.datapath")` so the
next map is instant. See the [Bathymetry article](bathymetry.html) for the full
list of sources and rendered examples.

# Build your own shapefiles

The create-your-own-bathymetry workflow is now complete. `raster_bathymetry()`
turns a GEBCO/ETOPO/IBCAO grid into a `bathyRaster`, and **two** vectorisers
consume it: `vector_bathymetry()` for depth-contour polygons (as before) and the
new `vector_land()` for a matching land polygon extracted from the same grid.

```{r}
rb <- raster_bathymetry(
  bathy = "path/to/your/grid.nc",
  depths = c(50, 100, 200, 500, 1000),
  proj.out = 4326,
  boundary = c(-5, 10, 50, 60)
)

basemap(
  limits = c(-5, 10, 50, 60),
  shapefiles = list(
    land   = vector_land(rb),       # NEW in v3
    glacier = NULL,
    bathy  = vector_bathymetry(rb)
  ),
  bathymetry = TRUE
)
```

Because land and bathymetry come from the same grid, their edges line up. The
[Customising shapefiles article](customising-shapefiles.html) walks through this
pipeline, `clip_shapefile()`, and reading Norwegian Geonorge depth data with
`geonorge_bathymetry()`.

# Expanded documentation

Version 3 splits the old single manual into a concise overview plus a set of
focused articles:

- **[Bathymetry](bathymetry.html)** — every way to get bathymetry into a map,
  from the shipped low-resolution grid to on-demand WCS and your own rasters.
- **[Customising shapefiles](customising-shapefiles.html)** — supplying your own
  land, glacier, and bathymetry layers.
- **[Adding graphical elements](adding-graphical-elements.html)** — ocean-current
  arrows (velocity quivers and schematic "Figure 1" arrows) and pie charts on
  maps via `scatterpie::geom_scatterpie()`.
- **[Cookbook](cookbook.html)** — short, copy-pasteable recipes.

# Tested and more reliable

Version 3 ships a comprehensive automated test suite: smoke tests covering the
historical regression corpus run everywhere, and `vdiffr` SVG snapshot tests
catch "code runs but wrong map" regressions. This release also fixes several
long-standing clipping problems. For example, projected maps with decimal-degree
limits used to cut off land near the map edges — `basemap(c(-20, 30, 50, 70))`
clipped off northern Norway. The clip boundary is now densified before
reprojection, so the full extent is kept:

```{r, eval = TRUE}
basemap(c(-20, 30, 50, 70), bathymetry = TRUE)
```

The map above is built entirely from the low-resolution data shipped with the
package, reprojected to Arctic stereographic on the fly — no downloads required.

# Upgrading from version 2

Version 3 is backwards compatible with version 2 map code; existing `basemap()`
and `qmap()` calls keep working. If you used the older bathymetry styles, see
the [version 2 release notes](new-features.html) and the
[Bathymetry article](bathymetry.html) for the current `bathy.style` names.
