---
title: "Pre-made shapefiles"
author: "Mikko Vihtakari (Institute of Marine Research)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{premade-shapefiles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(knitr)
knitr::opts_chunk$set(collapse = TRUE, message = FALSE, warning = FALSE, comment = "#>")

```

```{r setup}
library(ggOceanMaps)
```

Shapefiles used by this package have been defined as follows:

# Low-resolution shapefiles shipped with ggOceanMap

Download the datasets specified in [Data sources](#data-sources) to folders on your computer. Replace the paths under with the correct folder paths on your computer.

```{r}
etopoPath <- ""
NEDPath <- ""
outPath <- ""
```

## Decimal degrees

Bathymetry:

```{r eval = FALSE}
rb <- raster_bathymetry(bathy = paste(etopoPath, "ETOPO1_Ice_g_gmt4.grd", sep = "/"),
                        depths = c(50, 300, 500, 1000, 1500, 2000, 4000, 6000, 10000), 
                        proj.out = "+init=epsg:4326", 
                        boundary = c(-180.0083, 180.0083, -90, 90), 
                        aggregation.factor = 6
)

dd_bathy <- vector_bathymetry(rb, drop.crumbs = 50, remove.holes = 50)

save(dd_bathy, file = paste(outPath, "dd_bathy.rda", sep = "/"), compress = "xz")
```

Land:

```{r eval = FALSE}
world <- rgdal::readOGR(paste(NEDPath, "ne_10m_land/ne_10m_land.shp", sep = "/"))
islands <- rgdal::readOGR(paste(NEDPath, "ne_10m_minor_islands/ne_10m_minor_islands.shp", sep = "/"))
world <- rbind(world, islands)

dd_land <- clip_shapefile(world, c(-180, 180, -90, 90))

save(dd_land, file = paste(outPath, "dd_land.rda", sep = "/"), compress = "xz")
```

Glaciers:

```{r eval = FALSE}
glaciers <- rgdal::readOGR(paste(NEDPath, "ne_10m_glaciated_areas/ne_10m_glaciated_areas.shp", sep = "/"))
iceshelves <- rgdal::readOGR(paste(NEDPath, "ne_10m_antarctic_ice_shelves_polys/ne_10m_antarctic_ice_shelves_polys.shp", sep = "/"))

glaciers <- rbind(glaciers, iceshelves)
glaciers <- rgeos::gBuffer(glaciers, byid = TRUE, width = 0)

dd_glacier <- clip_shapefile(glaciers, c(-180, 180, -90, 90))
dd_glacier <- rgeos::gBuffer(dd_glacier, byid = FALSE, width = 0.1)
dd_glacier <- rgeos::gBuffer(dd_glacier, byid = FALSE, width = -0.1)

save(dd_glacier, file = paste(outPath, "dd_glacier.rda", sep = "/"), compress = "xz")
```

## Arctic stereographic

Bathymetry:

```{r eval = FALSE}
rb <- raster_bathymetry(bathy = paste(etopoPath, "ETOPO1_Ice_g_gmt4.grd", sep = "/"), 
                        depths = c(50, 300, 500, 1000, 1500, 2000, 4000, 6000, 10000), 
                        proj.out = "+init=epsg:3995", 
                        boundary = c(-180.0083, 180.0083, 30, 90), 
                        aggregation.factor = 2
)

arctic_bathy <- vector_bathymetry(rb)

save(arctic_bathy, file = paste(outPath, "arctic_bathy.rda", sep = "/"), compress = "xz")
```

Land:

```{r eval = FALSE}
arctic_land <- clip_shapefile(world, c(-180, 180, 30, 90))
arctic_land <- sp::spTransform(arctic_land, sp::CRS(sp::proj4string(arctic_bathy)))
arctic_land <- rgeos::gBuffer(arctic_land, byid = TRUE, width = 0)

save(arctic_land, file = paste(outPath, "arctic_land.rda", sep = "/"), compress = "xz")
```

Glaciers:

```{r eval = FALSE}
arctic_glacier <- clip_shapefile(glaciers, c(-180, 180, 40, 90))
arctic_glacier <- sp::spTransform(arctic_glacier, sp::CRS(sp::proj4string(arctic_bathy)))
arctic_glacier <- rgeos::gBuffer(arctic_glacier, byid = FALSE, width = 1000)
arctic_glacier <- rgeos::gBuffer(arctic_glacier, byid = FALSE, width = -1000)

save(arctic_glacier, file = paste(outPath, "arctic_glacier.rda", sep = "/"), compress = "xz")
```

## Antarctic steregraphic

Bathymetry:

```{r eval = FALSE}
rb <- raster_bathymetry(bathy = paste(etopoPath, "ETOPO1_Ice_g_gmt4.grd", sep = "/"), 
                        depths = c(50, 300, 500, 1000, 1500, 2000, 4000, 6000, 10000), 
                        proj.out = "+init=epsg:3031", 
                        boundary = c(-180.0083, 180.0083, -80, -30), 
                        aggregation.factor = 2
)

antarctic_bathy <- vector_bathymetry(rb)

save(antarctic_bathy, file = paste(outPath, "antarctic_bathy.rda", sep = "/"), compress = "xz")
```

Land:

```{r eval = FALSE}
antarctic_land <- clip_shapefile(world, c(-180, 180, -90, -30))
antarctic_land <- sp::spTransform(antarctic_land, sp::CRS(sp::proj4string(antarctic_bathy)))
antarctic_land <- rgeos::gBuffer(antarctic_land, byid = TRUE, width = 0)

save(antarctic_land, file = paste(outPath, "antarctic_land.rda", sep = "/"), compress = "xz")
```

Glaciers:

```{r eval = FALSE}
antarctic_glacier <- clip_shapefile(glaciers, c(-180, 180, -90, -30))
antarctic_glacier <- sp::spTransform(antarctic_glacier, sp::CRS(sp::proj4string(antarctic_bathy)))
antarctic_glacier <- rgeos::gBuffer(antarctic_glacier, byid = FALSE, width = 1000)
antarctic_glacier <- rgeos::gBuffer(antarctic_glacier, byid = FALSE, width = -1000)

save(antarctic_glacier, file = paste(outPath, "antarctic_glacier.rda", sep = "/"), compress = "xz")
```

# Data sources

The data used by the package are not the property of the Institute of Marine Research nor the author of the package. It is, therefore, important that you cite the data sources used in a map you generate with the package. The spatial data used by this package have been acquired from the following sources:

- **Land polygons.** [Natural Earth Data](https://www.naturalearthdata.com/downloads/10m-physical-vectors/) 1:10m Physical Vectors with the Land and Minor Island datasets combined. Distributed under the [CC Public Domain license](https://creativecommons.org/publicdomain/) ([terms of use](https://www.naturalearthdata.com/about/terms-of-use/)).
- **Glacier polygons.** [Natural Earth Data](https://www.naturalearthdata.com/downloads/10m-physical-vectors/) 1:10m Physical Vectors with the Glaciated Areas and Antarctic Ice Shelves datasets combined. Distributed under the [CC Public Domain license](https://creativecommons.org/publicdomain/) ([terms of use](https://www.naturalearthdata.com/about/terms-of-use/)).
- **Bathymetry.** [Amante, C. and B.W. Eakins, 2009. ETOPO1 1 Arc-Minute Global Relief Model: Procedures, Data Sources and Analysis. NOAA Technical Memorandum NESDIS NGDC-24. National Geophysical Data Center, NOAA](https://www.ngdc.noaa.gov/mgg/global/relief/ETOPO1/docs/ETOPO1.pdf). Distributed under the [U.S. Government Work license](https://www.usa.gov/government-works).

