---
title: "Pre-made shapefiles"
author: "Mikko Vihtakari (Institute of Marine Research)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{premade-shapefiles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(knitr)
knitr::opts_chunk$set(collapse = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      eval = FALSE,
                      comment = "#>"
)

```

```{r setup, eval = TRUE}
library(ggOceanMaps)
```

Shapefiles used by this package are stored in two places: 1) the low-resolution shapefiles used frequently are stored in the [ggOceanMapsData](https://github.com/MikkoVihtakari/ggOceanMapsData) package which should be installed automatically together with ggOceanMaps. 2) High-resolution shapefiles of selected locations are stored in the [ggOceanMapsLargeData](https://github.com/MikkoVihtakari/ggOceanMapsLargeData) repository and downloaded automatically as needed. This document describes how each of the shapefiles is created. 

# 1. Low-resolution shapefiles in ggOceanMapData

Download the datasets specified in [Data sources](#data-sources) to folders on your computer. Replace the paths under with the correct folder paths on your computer.

```{r eval = FALSE}
etopoPath <- ""
NEDPath <- ""
outPath <- ""
```

Define the the [coordinate reference systems](https://mgimond.github.io/Spatial/coordinate-systems-in-r.html).

```{r}
arcticCRS <- paste0("EPSG:", shapefile_list("Arctic")$crs)
antarcticCRS <- paste0("EPSG:", shapefile_list("Antarctic")$crs)
ddCRS <- paste0("EPSG:", shapefile_list("Decimal")$crs)
```

## Decimal degrees

Bathymetry:

```{r}
rb <- raster_bathymetry(
  bathy = file.path(etopoPath, "ETOPO1_Ice_g_gmt4.grd"),
  depths = c(50, 300, 500, 1000, 1500, 2000, 4000, 6000, 10000), 
  proj.out = ddCRS, 
  boundary = c(-180.0083, 180.0083, -90, 90), 
  aggregation.factor = 6
)

dd_bathy <- vector_bathymetry(rb, drop.crumbs = 50, remove.holes = 50)

save(dd_bathy, file = paste(outPath, "ggOceanMapsData/dd_bathy.rda", sep = "/"), compress = "xz")
```

Land:

```{r}
world <- sf::st_read(file.path(NEDPath, "ne_10m_land/ne_10m_land.shp"))
islands <- sf::st_read(file.path(NEDPath, "ne_10m_minor_islands/ne_10m_minor_islands.shp"))
world <- rbind(world, islands)

dd_land <- clip_shapefile(world, c(-180, 180, -90, 90))

save(dd_land, file = paste(outPath, "ggOceanMapsData/dd_land.rda", sep = "/"), compress = "xz")
```

Glaciers:

```{r}
glaciers <- sf::st_read(file.path(NEDPath, "ne_10m_glaciated_areas/ne_10m_glaciated_areas.shp"))
iceshelves <- sf::st_read(file.path(NEDPath, "ne_10m_antarctic_ice_shelves_polys/ne_10m_antarctic_ice_shelves_polys.shp"))

glaciers <- sf::as_Spatial(rbind(glaciers, iceshelves))
glaciers <- rgeos::gBuffer(glaciers, byid = TRUE, width = 0)

dd_glacier <- clip_shapefile(glaciers, c(-180, 180, -90, 90))
dd_glacier <- rgeos::gBuffer(dd_glacier, byid = FALSE, width = 0.1)
dd_glacier <- rgeos::gBuffer(dd_glacier, byid = FALSE, width = -0.1)

save(dd_glacier, file = paste(outPath, "ggOceanMapsData/dd_glacier.rda", sep = "/"), compress = "xz")
```

## Arctic stereographic

Bathymetry:

```{r}
rb <- raster_bathymetry(
  bathy = file.path(etopoPath, "ETOPO1_Ice_g_gmt4.grd"), 
  depths = c(50, 300, 500, 1000, 1500, 2000, 4000, 6000, 10000), 
  proj.out = arcticCRS, 
  boundary = c(-180.0083, 180.0083, 30, 90), 
  aggregation.factor = 2
)

arctic_bathy <- vector_bathymetry(rb)

save(arctic_bathy, file = paste(outPath, "ggOceanMapsData/arctic_bathy.rda", sep = "/"), compress = "xz")
```

Land:

```{r}
arctic_land <- clip_shapefile(world, c(-180, 180, 30, 90))
arctic_land <- sp::spTransform(arctic_land, sp::CRS(arcticCRS))
arctic_land <- rgeos::gBuffer(arctic_land, byid = TRUE, width = 0)

save(arctic_land, file = paste(outPath, "ggOceanMapsData/arctic_land.rda", sep = "/"), compress = "xz")
```

Glaciers:

```{r}
arctic_glacier <- clip_shapefile(glaciers, c(-180, 180, 40, 90))
arctic_glacier <- sp::spTransform(arctic_glacier, sp::CRS(arcticCRS))
arctic_glacier <- rgeos::gBuffer(arctic_glacier, byid = FALSE, width = 1000)
arctic_glacier <- rgeos::gBuffer(arctic_glacier, byid = FALSE, width = -1000)

save(arctic_glacier, file = paste(outPath, "ggOceanMapsData/arctic_glacier.rda", sep = "/"), compress = "xz")
```

## Antarctic steregraphic

Bathymetry:

```{r}
rb <- raster_bathymetry(
  bathy = file.path(etopoPath, "ETOPO1_Ice_g_gmt4.grd"), 
  depths = c(50, 300, 500, 1000, 1500, 2000, 4000, 6000, 10000), 
  proj.out = antarcticCRS, 
  boundary = c(-180.0083, 180.0083, -80, -30), 
  aggregation.factor = 2
)

antarctic_bathy <- vector_bathymetry(rb)

save(antarctic_bathy, file = paste(outPath, "ggOceanMapsData/antarctic_bathy.rda", sep = "/"), compress = "xz")
```

Land:

```{r}
antarctic_land <- clip_shapefile(world, c(-180, 180, -90, -30))
antarctic_land <- sp::spTransform(antarctic_land, sp::CRS(antarcticCRS))
antarctic_land <- rgeos::gBuffer(antarctic_land, byid = TRUE, width = 0)

save(antarctic_land, file = paste(outPath, "ggOceanMapsData/antarctic_land.rda", sep = "/"), compress = "xz")
```

Glaciers:

```{r}
antarctic_glacier <- clip_shapefile(glaciers, c(-180, 180, -90, -30))
antarctic_glacier <- sp::spTransform(antarctic_glacier, sp::CRS(antarcticCRS))
antarctic_glacier <- rgeos::gBuffer(antarctic_glacier, byid = FALSE, width = 1000)
antarctic_glacier <- rgeos::gBuffer(antarctic_glacier, byid = FALSE, width = -1000)

save(antarctic_glacier, file = paste(outPath, "ggOceanMapsData/antarctic_glacier.rda", sep = "/"), compress = "xz")
```

# 2. High-resolution shapefiles in ggOceanMapsLargeData

## Barents Sea

Bathymetry:

```{r}
bound.limits <- c(-25, 75, 65, 84)

rb <- raster_bathymetry(
  bathy = file.path(gebcoPath, "GEBCO_2020.nc"), 
  depths = c(50, 100, 200, 300, 400, 500, 750, 1000, 1500, 2000, 
             3000, 4000, 5000, 6000, 10000), 
  proj.out = paste0("EPSG:", shapefile_list("Barents")$crs), 
  boundary = bound.limits
)

barentssea_bathy <- vector_bathymetry(rb)

save(barentssea_bathy, 
     file = paste(outPath, "ggOceanMapsLargeData/barentssea_bathy.rda", sep = "/"), 
     compress = "xz"
)
```

Land:

```{r}
barentssea_land <- clip_shapefile(world, bound.limits)
barentssea_land <- sp::spTransform(
  barentssea_land, 
  sp::CRS(paste0("EPSG:", shapefile_list("Barents")$crs))
)
barentssea_land <- rgeos::gBuffer(barentssea_land, byid = TRUE, width = 0)

save(barentssea_land, 
     file = paste(outPath, "ggOceanMapsLargeData/barentssea_land.rda", sep = "/"),
     compress = "xz"
)
```

Glaciers:

```{r}
barentssea_glacier <- clip_shapefile(glaciers, bound.limits)
barentssea_glacier <- sp::spTransform(
  barentssea_glacier, 
  sp::CRS(paste0("EPSG:", shapefile_list("Barents")$crs))
)
barentssea_glacier <- rgeos::gBuffer(barentssea_glacier, byid = TRUE, width = 0)

save(barentssea_glacier, 
     file = paste(outPath, "ggOceanMapsLargeData/barentssea_glacier.rda", sep = "/"),
     compress = "xz"
)
```

## High resolution Arctic stereographic

```{r}
rb <- raster_bathymetry(
  bathy = "/Users/a22357/Downloads/IBCAO_v4_200m.nc", 
  depths = c(50, 100, 200, 300, 400, 500, 750, 1000, 1500, 2000, 
             3000, 4000, 5000, 6000, 10000), 
  proj.bathy = "EPSG:3996"
)

ibcao_bathy <- vector_bathymetry(rb)

save(ibcao_bathy, 
     file = paste(outPath, "ggOceanMapsLargeData/ibcao_bathy.rda", sep = "/"),
     compress = "xz"
)
```


# Data sources

The data used by the package are not the property of the Institute of Marine Research nor the author of the package. It is, therefore, important that you cite the data sources used in a map you generate with the package. Please see [here](https://mikkovihtakari.github.io/ggOceanMaps/#citations-and-data-sources) for a list of data sources. 